Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-08-26 paper.tex
Widdled down Section II a little bit

The strain $h$ observed in a GW detector from a wave emitted by a BNS system is related to the two polarizations of the wave and to the detector's antenna beam pattern response:
\begin{equation}
h=F_+ h_+ + F_\times h_\times.
\end{equation}
For a single interferometric detector, the antenna patterns take the form
\begin{eqnarray}
F_+(\theta,\phi,\psi)&=&-\frac{1}{2}(1+\cos^2\theta)\cos(2\phi)\cos(2\psi)\\
\nonumber
&&-\cos(\theta)\sin(2\phi)\sin(2\psi),\\
F_\times(\theta,\phi,\psi)&=&\frac{1}{2}(1+\cos^2\theta)\cos(2\phi)\sin(2\psi)\\
\nonumber
&&-\cos(\theta)\sin(2\phi)\cos(2\psi),
\end{eqnarray}
in a coordinate system where the two arms define the x-axis and the y-axis, and the z-axis is normal to the x-y plane \les{[cite gr-qc/0008066]}{}.  Here, $(\theta, \phi)$ are the sky position in spherical coordinates, and $\psi$ is the polarization angle \les{Should I define what the polarization angle is?}{}.  Assuming the binary components are point particles in a circular orbit and restricting to leading order in amplitude, the two polarizations of the emitted GW seen by an observer at a distance $D$ away are
\begin{eqnarray}
\label{eq:hoftp}
h_+(t)&=&-\frac{4\mu}{D}\left(\frac{1+\cos^2\iota}{2}\right)x(t)\cos\left[2\Phi(t)\right],\\
\label{eq:hoftc}
h_\times(t)&=&-\frac{4\mu}{D}\cos\iota\mbox{ }x(t)\sin\left[2\Phi(t)\right],
\end{eqnarray}
in the time-domain, where $\mu=m_1 m_2/M$ is the reduced mass, $M=m_1+m_2$ is the total mass, $\iota$ is the inclination angle \les{Should I define what the inclination angle is?}{}, $\Phi(t)$ is the orbital phase, and $x=(\pi M f)^{2/3}$ where $f$ is the GW frequency.  Using the Fourier transform convention
\begin{equation}
\tilde{h}(f)=\int_{-\infty}^{+\infty} h(t) e^{-2\pi i f t}dt,
\end{equation}
the stationary-phase approximation can be used to find the two polarizations in the frequency-domain:
\begin{eqnarray}
\label{eq:hoffp}
\tilde{h}_+(f)&=&\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\left(\frac{1+\cos^2\iota}{2}\right)e^{-i\Psi(f)},\\
\label{eq:hoffc}
\tilde{h}_\times(f)&=&-i\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\cos\iota\mbox{ }e^{-i\Psi(f)},
\end{eqnarray}
where the chirp mass ${\cal M}=\eta^{3/5}M$, the symmetric mass ratio $\eta=m_1 m_2/M^2$, $\Psi(f)=2\pi f t(v) - 2\Phi(v) - \pi/4$, and $\tilde{h}_+(f)$ and $\tilde{h}_\times(f)$ are the Fourier transforms of $h_+(t)$ and $h_\times(t)$ respectively.

The orbital phase evolution is described by the following system of ordinary differential equations:
\begin{eqnarray}
\label{eq:dPhidt}
\frac{d\Phi(t)}{dt}&=&\frac{x^{3/2}}{M},\\
\label{eq:dxdt}
\frac{d x(t)}{dt}&=&\frac{-{\cal L}}{E^{\prime}},
\end{eqnarray}
where $E^{\prime}=dE/dx$ is the analytic derivative of the energy $E$, and the luminosity ${\cal L}$ is related to energy via the energy balance requirement $dE/dt=-{\cal L}$.  Given an energy and luminosity, Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} can be solved to find $\Phi(t)$ and $x(t)$, which determines the two time-domain GW polarizations in Eqs.~\eqref{eq:hoftp} and~\eqref{eq:hoftc}.

Integrating Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} gives the alternate form:
\begin{eqnarray}
\label{eq:int_t}
t(x)&=&t_{\rm ref}+\int_x^{x_{\rm ref}}\frac{E^\prime(y)}{{\cal L}(y)}dy,\\
\label{eq:int_Phi}
\Phi(x)&=&\Phi_{\rm ref}+\int_x^{x_{\rm ref}}\frac{y^{3/2}}{M}\frac{E^\prime(y)}{{\cal L}(y)}dy,
\end{eqnarray}
where $t_{\rm ref}=t(x_{\rm ref})$, $\Phi_{\rm ref}=\Phi(x_{\rm ref})$, and $x_{\rm ref}$ is an arbitrary reference value, following \cite{BuonannoIyerOchsner2009}.  Given an energy and luminosity, Eqs.~\eqref{eq:int_t} and~\eqref{eq:int_Phi} can be solved and used to find $\Psi(f)$, which determines the two frequency-domain GW polarizations in Eqs.~\eqref{eq:hoffp} and~\eqref{eq:hoffc}.

@@

As was done in most previous works~\cite{WadeCreightonOchsner2014, DelPozzoLiAgathos2013, Favata2014, YagiYunes2014}, we keep the PN point-particle corrections to the energy and luminosity up to 3.5PN order as well as the leading-order and next-to-leading-order tidal contributions, which enter at the same order as 5PN and 6PN point-particle terms respectively.  The energy and luminosity take the form:
\begin{align}
E &= - \frac{1}{2} M \eta x \left[1 + e_\text{PP-PN}(\eta) + e_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right], \\
\mathcal{L} &= \frac{32}{5}\eta^2 x^5 \left[ 1 + l_\text{PP-PN}(\eta) + l_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right],
\end{align}
where the point-particle PN terms $e_\text{PP-PN}$ and $l_\text{PP-PN}$ are given in Ref.~\cite{Blanchet2014Review}, and the tidal terms are
\begin{widetext}
\begin{eqnarray}
\label{eq:e_tidal}
e_\text{Tidal} &=& -\frac{9\eta}{2} \left[ (1-3\eta)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}(1-\eta)(\Lambda_1-\Lambda_2) \right]x^5 \\
\nonumber
&& - 22\eta \left[ \left(1-4\eta+\frac{19\eta^2}{8}\right)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}\left(1-2\eta+\frac{3\eta^2}{8}\right)(\Lambda_1-\Lambda_2) \right]x^6,\\
\label{eq:l_tidal}
l_\text{Tidal}&=&3 \left[ (1-2\eta-4\eta^2)(\Lambda_1+\Lambda_2) + \sqrt{1-4\eta}(1-2\eta^2)(\Lambda_1-\Lambda_2) \right]x^5 \\
\nonumber
&& - \frac{22}{7} \left[ \left(1-\frac{15\eta}{176}+\frac{147\eta^2}{88}-\frac{1547\eta^3}{44}\right)(\Lambda_1+\Lambda_2) \right . \\
\nonumber
&& \left . + \sqrt{1-4\eta}\left(1+\frac{337\eta}{176}+\frac{15\eta^2}{2}-\frac{1085\eta^3}{88}\right)(\Lambda_1-\Lambda_2) \right]x^6,
\end{eqnarray}
\end{widetext}
where $\Lambda_1=\lambda_1/m_1$ and $\Lambda_2=\lambda_2/m_2$.  In order to write Eqs.~\eqref{eq:e_tidal} and~\eqref{eq:l_tidal} in this form, we have also asserted that $m_1 \ge m_2$ in order to express each correction in terms of dimensionless quantities.

As is described in detail in the Appendix of Ref.~\cite{WadeCreightonOchsner2014}, there are several methods to determine the phase evolution of the system from its energy and luminosity, and we use three of them.  In the TaylorT1 method, one numerically integrates Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} to find $\Phi(t)$ and $x(t)$. The TaylorT4 method is a slight variation where one re-expands the ratio $-{\cal L}/E^\prime$ on right hand side of Eq.~\eqref{eq:dxdt} then truncates the series at the highest known PN order (here 3.5PN) before numerically integrating to find $\Phi(t)$ and $x(t)$. The TaylorF2 method uses the stationary phase approximation to estimate the Fourier transformed waveform.  Eqs.~\eqref{eq:int_t} and~\eqref{eq:int_Phi} are used to write $\Psi(f)$ in the following form:
\begin{equation}
\label{eq:psi}
\Psi(f)=2\pi f t_{\rm ref} - 2\Phi_{\rm ref} -\frac{\pi}{4}+ 2\int_x^{x_{\rm ref}}\frac{x^{3/2} - y^{3/2}}{M}\frac{E^\prime(y)}{{\cal L}(y)}dy.
\end{equation}
The ratio $E^\prime/{\cal L}$ is then re-expanded and truncated at the highest known PN order (here 3.5PN) before the integral in Eq.~\eqref{eq:psi} is analytically computed to find $\Psi(f)$.  The differences in the PN waveform families are due mostly to the way terms beyond 3.5PN order are truncated.  Thus, the differences in results using one waveform family as the injection and another as templates should give a reasonable estimate of systematic errors due to unknown higher order terms.

The majority of this work uses the TaylorF2 waveform approximant because its fully analytic frequency-domain form makes it the fastest PN waveform to generate.  The complete TaylorF2 strain can be written more compactly in the following form:
\begin{equation}
\tilde{h}(f)=\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}}{\pi^{2/3}}\frac{Q(\theta,\phi,\psi,\iota)}{D}f^{-7/6}e^{-i\Psi(f)},
\end{equation}
where
\begin{equation}
Q(\theta,\phi,\psi,\iota) = F_+(\theta,\phi,\psi)\frac{1+\cos^2\iota}{2} - i F_\times(\theta,\phi,\psi)\cos\iota.
\end{equation}
For a single detector, the extrinsic parameters in the amplitude can be grouped together into a single effective distance parameter, which is $D_{\rm eff}=D/|Q(\theta,\phi,\psi,\iota)|$.  The analytic form for the phase $\Psi(f)$ with point-particle terms up to 3.5PN order and leading-order and next-to-leading-order tidal contributions is
\begin{eqnarray}
\Psi(f)&=&2\pi f t_{\rm ref} - 2\Phi_{\rm ref} -\frac{\pi}{4}\\
\nonumber
&&+\frac{3}{128\eta x^{5/2}}\left[1+\psi_\text{PP-PN}(\eta)+\psi_\text{Tidal}(\eta,\Lambda_1,\Lambda_2)\right],
\end{eqnarray}
where the point-particle PN terms $\psi_\text{PP-PN}(\eta)$ are given in \cite{BuonannoIyerOchsner2009}, and
\begin{widetext}
\begin{equation}
\label{eq:psi_tidal}
\psi_\text{Tidal}(\eta,\Lambda_1,\Lambda_2)=\left[-\frac{39}{2}\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)\right]x^5+ \left[-\frac{3115}{64}\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)+\frac{6595}{364}\sqrt{1-4\eta}\mbox{ }\delta\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)\right]x^6,
\end{equation}
where
\begin{eqnarray}
\label{eq:LT}
\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)&=&\frac{8}{13}\left[\left(1+7\eta-31\eta^2\right)\left(\Lambda_1+\Lambda_2\right)+\sqrt{1-4\eta}\left(1+9\eta-11\eta^2\right)\left(\Lambda_1-\Lambda_2\right)\right],\\
\label{eq:dLT}
\delta\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)&=&\frac{1}{2}\left[\sqrt{1-4\eta}\left(1-\frac{13272}{1319}\eta+\frac{8944}{1319}\eta^2\right)\left(\Lambda_1+\Lambda_2\right) \right .\\
&& + \left . \left(1-\frac{15910}{1319}\eta+\frac{32850}{1319}\eta^2+\frac{3380}{1319}\eta^3\right)\left(\Lambda_1-\Lambda_2\right)\right].
\end{eqnarray}
\end{widetext}
We express the tidal contributions to the TaylorF2 phase in terms of $(\tilde\Lambda,\delta\tilde\Lambda)$ because $\tilde\Lambda$ is comparably more measurable than $\Lambda_1$ and $\Lambda_2$ individually.  For more details regarding this parameterization, see \cite{WadeCreightonOchsner2014,Favata2014}.

>>>>

The strain $h$ observed in a GW detector from a wave emitted by a BNS system is related to the two polarizations of the wave ($h_+$ and $h_\times$) and to the detector's antenna beam pattern response ($F_+$ and $F_\times$).  Assuming the binary components are point particles in a circular orbit and restricting to leading order in amplitude, the two polarizations of the emitted GW seen by an observer at a distance $D$ away are
\begin{eqnarray}
\label{eq:hoftp}
h_+(t)&=&-\frac{4\mu}{D}\left(\frac{1+\cos^2\iota}{2}\right)x(t)\cos\left[2\Phi(t)\right],\\
\label{eq:hoftc}
h_\times(t)&=&-\frac{4\mu}{D}\cos\iota\mbox{ }x(t)\sin\left[2\Phi(t)\right],
\end{eqnarray}
in the time-domain, where $\mu=m_1 m_2/M$ is the reduced mass, $M=m_1+m_2$ is the total mass, $\iota$ is the inclination angle \les{Should I define what the inclination angle is?}{}, $\Phi(t)$ is the orbital phase, and $x=(\pi M f)^{2/3}$ where $f$ is the GW frequency.  Using the Fourier transform convention
\begin{equation}
\tilde{h}(f)=\int_{-\infty}^{+\infty} h(t) e^{-2\pi i f t}dt,
\end{equation}
the stationary-phase approximation can be used to find the two polarizations in the frequency-domain:
\begin{eqnarray}
\label{eq:hoffp}
\tilde{h}_+(f)&=&\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\left(\frac{1+\cos^2\iota}{2}\right)e^{-i\Psi(f)},\\
\label{eq:hoffc}
\tilde{h}_\times(f)&=&-i\sqrt{\frac{5}{24}}\frac{{\cal M}^{5/6}f^{-7/6}}{\pi^{2/3}D}\cos\iota\mbox{ }e^{-i\Psi(f)},
\end{eqnarray}
where the chirp mass ${\cal M}=\eta^{3/5}M$, the symmetric mass ratio $\eta=m_1 m_2/M^2$, $\Psi(f)=2\pi f t(v) - 2\Phi(v) - \pi/4$, and $\tilde{h}_+(f)$ and $\tilde{h}_\times(f)$ are the Fourier transforms of $h_+(t)$ and $h_\times(t)$ respectively.  Given an energy $E$ and luminosity ${\cal L}$, $\Phi(t)$ and $x(t)$, which determines the two time-domain GW polarizations in Eqs.~\eqref{eq:hoftp} and~\eqref{eq:hoftc}, and $\Psi(f)$, which determines the two frequency-domain GW polarizations in Eqs.~\eqref{eq:hoffp} and~\eqref{eq:hoffc}, can be found from the following ordinary differential equations:
\begin{eqnarray}
\label{eq:dPhidt}
\frac{d\Phi(t)}{dt}&=&\frac{x^{3/2}}{M},\\
\label{eq:dxdt}
\frac{d x(t)}{dt}&=&\frac{-{\cal L}}{E^{\prime}},
\end{eqnarray}
where $E^{\prime}=dE/dx$ is the analytic derivative of the energy $E$, and the luminosity ${\cal L}$ is related to energy via the energy balance requirement $dE/dt=-{\cal L}$.

@@

As was done in most previous works~\cite{WadeCreightonOchsner2014, DelPozzoLiAgathos2013, Favata2014, YagiYunes2014}, we keep the PN point-particle corrections to the energy and luminosity up to 3.5PN order as well as the leading-order and next-to-leading-order tidal contributions, which enter at the same order as 5PN and 6PN point-particle terms respectively.  The energy and luminosity take the form:
\begin{align}
E &= - \frac{1}{2} M \eta x \left[1 + e_\text{PP-PN}(\eta) + e_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right], \\
\mathcal{L} &= \frac{32}{5}\eta^2 x^5 \left[ 1 + l_\text{PP-PN}(\eta) + l_\text{Tidal}(\eta, \Lambda_1, \Lambda_2)\right],
\end{align}
where the point-particle PN terms $e_\text{PP-PN}$ and $l_\text{PP-PN}$ are given in Ref.~\cite{Blanchet2014Review}, and the tidal terms are given in Ref.~\cite{WadeCreightonOchsner2014}.  In order to write $e_\text{Tidal}$ and $l_\text{Tidal}$ as a function of the dimensionless quantity $\eta$, we have asserted that $m_1 \ge m_2$.

As is described in detail in the Appendix of Ref.~\cite{WadeCreightonOchsner2014}, there are several methods to determine the phase evolution of the system from its energy and luminosity, and we use three of them.  In the TaylorT1 method, one numerically integrates Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} to find $\Phi(t)$ and $x(t)$. The TaylorT4 method is a slight variation where one re-expands the ratio $-{\cal L}/E^\prime$ on right hand side of Eq.~\eqref{eq:dxdt} then truncates the series at the highest known PN order (here 3.5PN) before numerically integrating to find $\Phi(t)$ and $x(t)$. The TaylorF2 method uses the stationary phase approximation to estimate the Fourier transformed waveform.  Eqs.~\eqref{eq:dPhidt} and~\eqref{eq:dxdt} are analytically integrated to write $\Psi(f)$ in the following form:
\begin{equation}
\label{eq:psi}
\Psi(f)=2\pi f t_{\rm ref} - 2\Phi_{\rm ref} -\frac{\pi}{4}+ 2\int_x^{x_{\rm ref}}\frac{x^{3/2} - y^{3/2}}{M}\frac{E^\prime(y)}{{\cal L}(y)}dy,
\end{equation}
where $t_{\rm ref}=t(x_{\rm ref})$, $\Phi_{\rm ref}=\Phi(x_{\rm ref})$, and $x_{\rm ref}$ is an arbitrary reference value, following \cite{BuonannoIyerOchsner2009}.  The ratio $E^\prime/{\cal L}$ is then re-expanded and truncated at the highest known PN order (here 3.5PN) before the integral in Eq.~\eqref{eq:psi} is analytically computed to find $\Psi(f)$.  The differences in the PN waveform families are due mostly to the way terms beyond 3.5PN order are truncated.  Thus, the differences in results using one waveform family as the injection and another as templates should give a reasonable estimate of systematic errors due to unknown higher order terms.

The majority of this work uses the TaylorF2 waveform approximant because its fully analytic frequency-domain form makes it the fastest PN waveform to generate.  The form of the phase $\Psi(f)$ with point-particle terms up to 3.5PN order and leading-order and next-to-leading-order tidal contributions is
\begin{eqnarray}
\Psi(f)&=&2\pi f t_{\rm ref} - 2\Phi_{\rm ref} -\frac{\pi}{4}\\
\nonumber
&&+\frac{3}{128\eta x^{5/2}}\left[1+\psi_\text{PP-PN}(\eta)+\psi_\text{Tidal}(\eta,\Lambda_1,\Lambda_2)\right],
\end{eqnarray}
where the point-particle PN terms $\psi_\text{PP-PN}(\eta)$ are given in \cite{BuonannoIyerOchsner2009}, and
\begin{widetext}
\begin{equation}
\label{eq:psi_tidal}
\psi_\text{Tidal}(\eta,\Lambda_1,\Lambda_2)=\left[-\frac{39}{2}\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)\right]x^5+ \left[-\frac{3115}{64}\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)+\frac{6595}{364}\sqrt{1-4\eta}\mbox{ }\delta\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)\right]x^6,
\end{equation}
where $\Lambda_1=\lambda_1/m_1^5$, $\Lambda_2=\lambda_2/m_2^5$, and
\begin{eqnarray}
\label{eq:LT}
\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)&=&\frac{8}{13}\left[\left(1+7\eta-31\eta^2\right)\left(\Lambda_1+\Lambda_2\right)+\sqrt{1-4\eta}\left(1+9\eta-11\eta^2\right)\left(\Lambda_1-\Lambda_2\right)\right],\\
\label{eq:dLT}
\delta\tilde{\Lambda}(\eta,\Lambda_1,\Lambda_2)&=&\frac{1}{2}\left[\sqrt{1-4\eta}\left(1-\frac{13272}{1319}\eta+\frac{8944}{1319}\eta^2\right)\left(\Lambda_1+\Lambda_2\right) \right .\\
&& + \left . \left(1-\frac{15910}{1319}\eta+\frac{32850}{1319}\eta^2+\frac{3380}{1319}\eta^3\right)\left(\Lambda_1-\Lambda_2\right)\right].
\end{eqnarray}
\end{widetext}
We express the tidal contributions to the TaylorF2 phase in terms of $(\tilde\Lambda,\delta\tilde\Lambda)$ because $\tilde\Lambda$ is comparably more measurable than $\Lambda_1$ and $\Lambda_2$ individually.  For more details regarding this parameterization, see \cite{WadeCreightonOchsner2014,Favata2014}.
